Uncertainty propagation in deep neural networks for sparse coding

### 2018-03-07 paper/BayesianLISTA.tex
reorganisation + collect results on synthetic + paper correction

\title{Bayesian LISTA / Uncertainty propagation in deep neural networks for sparse coding}

@@

\begin{lemma}[Moments of spike and slab]
\label{thm:moments_spsl}
Let the random variable $\xi$ have a spike and slab distribution with probability of spike $\omega$, slab mean $m$ and slab variance $v$. Then its moments are
\begin{align}
\mathbb{E}\xi &= (1-\omega)m \\
%\mathbb{E}\xi^2 &= (1-\omega)(v + m^2) \\
\text{Var }\xi & = (1-\omega)(v + \omega m^2)
\end{align}

\end{lemma}
\begin{proof}
\begin{align}
\mathbb{E}\xi &= \int x (\omega \delta_0(x) + (1 - \omega)\mathcal{N}(x; m, v))dx \\
& = \omega \int x \delta_0(x)dx + (1 - \omega)\int x \mathcal{N}(x; m, v)dx \\
& = (1-\omega)m
\end{align}
\begin{align}
\mathbb{E}\xi^2 &= \int x^2 (\omega \delta_0(x) + (1 - \omega)\mathcal{N}(x; m, v))dx \\
& = \omega \int x^2 \delta_0(x)dx + (1 - \omega)\int x^2 \mathcal{N}(x; m, v)dx \\
& = (1-\omega)(v + m^2)
\end{align}
\begin{align}
\text{Var }\xi &= \mathbb{E}\xi^2 - \left(\mathbb{E}\xi\right)^2\\
& = (1-\omega)(v + \omega m^2)
\end{align}
\end{proof}

 \begin{lemma}
 \label{thm:matrix_const}
 Let $\mathbf{W} \in \mathbb{R}^{D \times K}$ be a matrix of independent Gaussian-distributed random variables: $w_{dk} \sim \mathcal{N}(m^w_{dk}, v^w_{dk})$, and $\mathbf{y} \in \mathbb{R}^K$ be a constant vector. Then their product $\mathbf{W} \mathbf{y}$ is a vector $\mathbf{b} \in \mathbb{R}^{D}$ of random variables $b_d \sim \mathcal{N}(m^b_d, w^b_d)$, where $m^b_d = \sum_{k=1}^Ky_k m^w_{dk}$, $w^b_d = \sum_{k=1}^Ky_k^2v^w_{dk}$. 
 %If we denote $M^W = [m^w_{dk}]_{d=1:D, k=1:K}$, $V^W = [v^w_{dk}]_{d=1:D, k=1:K}$, then $\mathbf{b}^m = W^My$, $\mathbf{b}^w = W^Vy^2$.
 \end{lemma}
 \begin{proof}
 	The statement follows from the property that the family of normal distributions is closed under linear transformations.
 \end{proof}

  \begin{lemma}
  \label{thm:matrix_vector}
 Let $\mathbf{S} \in \mathbb{R}^{D \times D}$ be a matrix of independent Gaussian-distributed random variables: $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\boldsymbol\beta \in \mathbb{R}^D$ be a vector with spike-and-slab distributed variables: $\beta_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$. Then their product can be approximated as a vector $\mathbf{d} \in \mathbb{R}^{D}$ of random variables $d_d \sim \mathcal{N}(d^d_m, d^d_w)$, where $m^d_d = \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}$, $v^d_d = \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} + (1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}]$.
 \end{lemma}
 \begin{proof}
 	We compute the mean and variance of the product $\mathbf{S}\boldsymbol\beta$ and approximate resulting distribution as Gaussian.
\begin{align}
	&\mathbb{E}d_d = \mathbb{E} \left[\sum_{d'=1}^D s_{dd'}\beta_{d'}\right] =  \sum_{d'=1}^D \mathbb{E}[s_{dd'}\beta_{d'}] \\
	&= \sum_{d'=1}^D \mathbb{E}s_{dd'}\mathbb{E}\beta_{d'} = \sum_{d'=1}^D m^s_{dd'}\mathbb{E}\beta_{d'}
\end{align}
\begin{align}
	&\text{Var}d_d = \text{Var} \left[\sum_{d'=1}^D s_{dd'}\beta_{d'}\right] =  \sum_{d'=1}^D \text{Var}[s_{dd'}\beta_{d'}] = \\
	&\sum_{d'=1}^D [(\mathbb{E}s_{dd'})^2 \text{Var}\beta_{d'} + (\mathbb{E}\beta_{d'})^2 \text{Var}s_{dd'} + \text{Var}\beta_{d'} \text{Var}s_{dd'}]
\end{align}
where $\mathbb{E}\beta_{d'}$, $\text{Var}\beta_{d'}$ are computed according to lemma~\ref{thm:moments_spsl}.
 \end{proof}

\begin{lemma}
\label{thm:sum_vectors}
If $\mathbf{b} \in \mathbb{R}^{D}$ and $\mathbf{d} \in \mathbb{R}^{D}$ are both vectors of independent Gaussian-distributed random variables: $b_{d} \sim \mathcal{N}(m^b_{d}, v^b_{d})$, $d_{d} \sim \mathcal{N}(m^d_{d}, v^d_{d})$ then their sum $\mathbf{c} = \mathbf{b} + \mathbf{d}$ is a vector of independent Gaussian-distributed random variables $c_{d} \sim \mathcal{N}(m^c_{d}, v^c_{d})$ with $m^c_{d} = m^b_{d} + m^d_{d}$, $v^c_{d} = v^b_{d} + v^d_{d}$
\end{lemma}
\begin{proof}
Based on properties of Gaussian distributions.
\end{proof}

\begin{lemma}
\label{thm:soft_thresholding}
When a Gaussian-distributed random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft-thresholding function it can be approximated with the spike and slab distribution on $x^*$ with the probability of spike $\omega^*$, slab mean $m^*$ and slab variance $v^*$ (Formulae are given in the proof).
\end{lemma}
\begin{proof}
Soft-thresholding operator is defined as $h_\lambda(v) = \text{sign}(v) * \max(|v| - \lambda, 0)$. We need to compute the parameters of approximating distribution.
\begin{enumerate}
\item The probability of spike $\omega^*$ equals to the probability mass that lies in $(-\lambda, \lambda)$ of the original distribution and is flattened into zero by soft-thresholding operator. As the original distribution is Gaussian, this can be computed as $\omega^* = p(x^*=0) = p(x \in (-\lambda, \lambda)) = \Phi(\frac{\lambda-m}{\sqrt{v}}) - \Phi(\frac{-\lambda-m}{\sqrt{v}})$. 
\item Soft-thresholding shifts elements that are greater than $\lambda$  or less than $-\lambda$ elements towards 0. Let $\psi(\cdot)$ denote the density of the soft-threshold distribution, $\phi(\cdot)$ denote the density of the original Gaussian distribution. Then the first moment of the resulting distribution is $m^* = \int_{-\infty}^{+\infty}x\psi(x)dx = \int_{-\infty}^{0}x\phi(x-\lambda)dx + \int_{0}^{+\infty}x\phi(x+\lambda)dx$. Integrals are computed below.
\begin{align}
&\int_{-\infty}^{0}x\phi(x-\lambda)dx = \\
&-\frac{\sqrt{v}}{\sqrt{2\pi}} \exp\left\{\frac{-(\lambda+m)^2}{2v}\right\} \\
& + (\lambda+m)\Phi\left(-\frac{\lambda+m}{\sqrt{v}}\right)
\end{align}
\begin{align}
&\int_{0}^{+\infty}x\phi(x+\lambda)dx = \\
& = \frac{\sqrt{v}}{\sqrt{2\pi}} \exp\left\{\frac{-(m - \lambda)^2}{2v}\right\}\\
& + (m - \lambda)\left(1 - \Phi\left(-\frac{\lambda-m}{\sqrt{v}}\right)\right)
\end{align}
\item The second moment of the approximating distribution is computed as $s = \int_{-\infty}^{+\infty}x^2\psi(x)dx = \int_{-\infty}^{0}x^2\phi(x-\lambda)dx + \int_{0}^{+\infty}x^2\phi(x+\lambda)dx$.
\begin{align}
&\int_{-\infty}^{0}x^2\phi(x-\lambda)dx = \\
&-\frac{\sqrt{v}}{\sqrt{2\pi}} (\lambda+m)\exp\left\{\frac{-(\lambda+m)^2}{2v}\right\}\\
& + (\sigma^2 + (\lambda+m)^2)\Phi\left(-\frac{\lambda+m}{\sqrt{v}}\right)
\end{align}
\begin{align}
&\int_{0}^{+\infty}x^2\phi(x+\lambda)dx = \\
&\frac{\sqrt{v}}{\sqrt{2\pi}} (m - \lambda)\exp\left\{\frac{-(m - \lambda)^2}{2v}\right\}\\
& + (\sigma^2 + (m - \lambda)^2)\left(1 - \Phi\left(\frac{\lambda -m}{\sqrt{v}}\right)\right)
\end{align}
Resulting variance is
\begin{equation}
v^* = s- (m^*)^2
\end{equation}
\end{enumerate}
\end{proof}

The proposed operations on Gaussian and spike-and-slab distributions allow to formulate the main result in this section that describes how the uncertainty propagation works.
\begin{theorem}[Bayesian LISTA forward propagation]
\label{thm:prob_layer}
The spike and slab distribution for $\boldsymbol\beta_{l-1}$ with parameters ($\boldsymbol\omega$, $\mathbf{m}$, $\mathbf{w}$) can be propagated through the LISTA layer and the parameters after propagation can be computed as following:
\begin{enumerate}
	\item $\mathbf{b} = Wy$ is computed according to lemma \ref{thm:matrix_const}
	\item $\mathbf{d} = S\boldsymbol\beta_{l-1}$ is computed according to lemma \ref{thm:matrix_vector}
	\item $\mathbf{c} = \mathbf{b} + \mathbf{d}$ is computed according to lemma \ref{thm:sum_vectors}
	\item $\boldsymbol\beta_{l} = h_\lambda(\mathbf{c})$ is computed according to lemma \ref{thm:soft_thresholding}
\end{enumerate}
\end{theorem}

\subsection{Approximation quality}
We have used two approximations in forward propagation of the uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft-thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.

>>>>

\title{Uncertainty propagation in deep neural networks for sparse coding}

@@

\begin{lemma}[Moments of spike and slab distribution]
\label{thm:moments_spsl}
Let the random variable $\xi$ have a spike and slab distribution with probability of spike $\omega$, slab mean $m$ and slab variance $v$. Then its moments are
\begin{subequations}
\begin{align}
\mathbb{E}\xi &= (1-\omega)m \\
\operatorname{Var}\xi & = (1-\omega)(v + \omega m^2)
\end{align}
\end{subequations}
\end{lemma}

\begin{proof}
\begin{align*}
\begin{split}
\mathbb{E}\xi &= \int x (\omega \delta_0(x) + (1 - \omega)\mathcal{N}(x; m, v))dx \\
& = \omega \int x \delta_0(x)dx + (1 - \omega)\int x \mathcal{N}(x; m, v)dx \\
& = (1-\omega)m \\
\mathbb{E}\xi^2 &= \int x^2 (\omega \delta_0(x) + (1 - \omega)\mathcal{N}(x; m, v))dx \\
& = \omega \int x^2 \delta_0(x)dx + (1 - \omega)\int x^2 \mathcal{N}(x; m, v)dx \\
& = (1-\omega)(v + m^2) \\
\operatorname{Var}\xi &= \mathbb{E}\xi^2 - \left(\mathbb{E}\xi\right)^2 = (1-\omega)(v + \omega m^2)
\end{split}
\end{align*}
\end{proof}

 \begin{lemma}[Product of Gaussian matrix and constant vector]
 \label{thm:matrix_const}
 Let $\mathbf{W} \in \mathbb{R}^{D \times K}$ be a matrix of independent Gaussian-distributed random variables: $w_{dk} \sim \mathcal{N}(m^w_{dk}, v^w_{dk})$, and $\mathbf{y} \in \mathbb{R}^K$ be a constant vector. Then their product $\mathbf{W} \mathbf{y}$ is a vector $\mathbf{b} \in \mathbb{R}^{D}$ of random variables $b_d \sim \mathcal{N}(m^b_d, w^b_d)$, where 
\begin{subequations}
\begin{align}
m^b_d &= \sum_{k=1}^Ky_k m^w_{dk}, \\
w^b_d &= \sum_{k=1}^Ky_k^2v^w_{dk}.
 \end{align}
\end{subequations}

 \end{lemma}
 \begin{proof}
 	The statement follows from the property that the family of normal distributions is closed under linear transformations.
 \end{proof}

  \begin{lemma}[Product of Gaussian matrix and Gaussian vector]
  \label{thm:matrix_vector}
 Let $\mathbf{S} \in \mathbb{R}^{D \times D}$ be a matrix of independent Gaussian-distributed random variables: $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\boldsymbol\beta \in \mathbb{R}^D$ be a vector with spike-and-slab distributed variables: $\beta_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$. If their product is approximated as a vector $\mathbf{e} \in \mathbb{R}^{D}$ of random variables $e_d \sim \mathcal{N}(m^e_d, w^e_d)$, then 
\begin{subequations}
\begin{align}
 m^e_d &= \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}, \\
 \begin{split}
 v^e_d &= \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} \\
 & {}+ (1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}].
 \end{split}
 \end{align}
\end{subequations}
 \end{lemma}
 \begin{proof}
 	We compute the mean and variance of the product $\mathbf{S}\boldsymbol\beta$ with an assumption of independence of its components and approximate resulting distribution with Gaussian. The quality of this approximation is discussed in section \ref{sec:approx_quality}
\begin{flalign*}
	\mathbb{E}e_d &= \sum_{d'=1}^D \mathbb{E}[s_{dd'}\beta_{d'}]  = \sum_{d'=1}^D m^s_{dd'}\mathbb{E}\beta_{d'}\\
	\operatorname{Var}e_d &= \sum_{d'=1}^D \operatorname{Var}[s_{dd'}\beta_{d'}] = \sum_{d'=1}^D [(\mathbb{E}s_{dd'})^2 \operatorname{Var}\beta_{d'} \\
	&{}+ (\mathbb{E}\beta_{d'})^2 \operatorname{Var}s_{dd'} + \operatorname{Var}\beta_{d'} \operatorname{Var}s_{dd'}]
\end{flalign*}
where $\mathbb{E}\beta_{d'}$, $\operatorname{Var}\beta_{d'}$ are computed according to lemma~\ref{thm:moments_spsl}.
 \end{proof}

\begin{lemma}[Sum of Gaussian vectors]
\label{thm:sum_vectors}
If $\mathbf{b} \in \mathbb{R}^{D}$ and $\mathbf{e} \in \mathbb{R}^{D}$ are both vectors of independent Gaussian-distributed random variables: $b_{d} \sim \mathcal{N}(m^b_{d}, v^b_{d})$, $e_{d} \sim \mathcal{N}(m^e_{d}, v^e_{d})$ then their sum $\mathbf{c} = \mathbf{b} + \mathbf{e}$ is a vector of independent Gaussian-distributed random variables $c_{d} \sim \mathcal{N}(m^c_{d}, v^c_{d})$ with 
\begin{subequations}
\begin{align}
m^c_{d} &= m^b_{d} + m^e_{d}, \\
v^c_{d} &= v^b_{d} + v^e_{d}.
 \end{align}
\end{subequations}
\end{lemma}
\begin{proof}
Based on properties of Gaussian distributions.
\end{proof}

\begin{lemma}[Gaussian propagation through soft thresholding]
\label{thm:soft_thresholding}
When a Gaussian-distributed random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft-thresholding function it can be approximated with the spike and slab distribution on $x^*$ with the probability of spike $\omega^*$, slab mean $m^*$ and slab variance $v^*$ (parameters of the approximating distribution are given in the proof).
\end{lemma}
\begin{proof}

The probability of spike $\omega^*$ equals to the probability mass of the original distribution that lies in $(-\lambda, \lambda)$  and is flattened into zero by soft-thresholding operator. As the original distribution is Gaussian, this can be computed as 
\begin{align}
\begin{split}
\omega^* &= p(x^*=0) = p(x \in [-\lambda, \lambda]) \\
&= \Phi\left(\frac{\lambda-m}{\sqrt{v}}\right) - \Phi\left(\frac{-\lambda-m}{\sqrt{v}}\right). 
\end{split}
\end{align}

Soft-thresholding shifts elements that are greater than $\lambda$  or less than $-\lambda$ elements towards 0. Let $\psi(\cdot)$ denote the density of the soft-threshold distribution, $\phi(\cdot)$ denote the density of the original Gaussian distribution. Then the first moment of the resulting distribution is

\begin{align}
\begin{split}
m^* &= \int_{-\infty}^{+\infty}x\psi(x)dx= \int_{-\infty}^{0}x\phi(x-\lambda)dx \\
&{} + \int_{0}^{+\infty}x\phi(x+\lambda)dx.
\end{split}
\end{align}

Integrals are computed as
\begin{align}
\begin{split}
&\int_{-\infty}^{0}x\phi(x-\lambda)dx = -\frac{\sqrt{v}}{\sqrt{2\pi}} \exp\left\{\frac{-(\lambda+m)^2}{2v}\right\} \\
&{} + (\lambda+m)\Phi\left(-\frac{\lambda+m}{\sqrt{v}}\right)
\end{split}\\
\begin{split}
&\int_{0}^{+\infty}x\phi(x+\lambda)dx = \frac{\sqrt{v}}{\sqrt{2\pi}} \exp\left\{\frac{-(m - \lambda)^2}{2v}\right\}\\
& + (m - \lambda)\left(1 - \Phi\left(-\frac{\lambda-m}{\sqrt{v}}\right)\right)
\end{split}
\end{align}

The second moment of the approximating distribution is computed as
\begin{align}
\begin{split}
s &= \int_{-\infty}^{+\infty}x^2\psi(x)dx = \int_{-\infty}^{0}x^2\phi(x-\lambda)dx \\
&{}+ \int_{0}^{+\infty}x^2\phi(x+\lambda)dx.
\end{split}
\end{align}

Integrals are computed as
\begin{align}
\begin{split}
&\int_{-\infty}^{0}x^2\phi(x-\lambda)dx = \\
&-\frac{\sqrt{v}}{\sqrt{2\pi}} (\lambda+m)\exp\left\{\frac{-(\lambda+m)^2}{2v}\right\}\\
& + (\sigma^2 + (\lambda+m)^2)\Phi\left(-\frac{\lambda+m}{\sqrt{v}}\right)
\end{split}\\
\begin{split}
&\int_{0}^{+\infty}x^2\phi(x+\lambda)dx = \\
&\frac{\sqrt{v}}{\sqrt{2\pi}} (m - \lambda)\exp\left\{\frac{-(m - \lambda)^2}{2v}\right\}\\
& + (\sigma^2 + (m - \lambda)^2)\left(1 - \Phi\left(\frac{\lambda -m}{\sqrt{v}}\right)\right)
\end{split}
\end{align}

Resulting variance is
\begin{equation}
v^* = s- (m^*)^2
\end{equation}

\end{proof}

The proposed operations on Gaussian and spike-and-slab distributions allow to formulate the main result in this section that describes how the uncertainty propagation works.
\begin{theorem}[Bayesian LISTA forward propagation]
\label{thm:prob_layer}
The spike and slab distribution for $\boldsymbol\beta_{l-1}$ with parameters ($\boldsymbol\omega$, $\mathbf{m}$, $\mathbf{w}$) can be propagated through the LISTA layer and the parameters after propagation can be computed as following:
\begin{enumerate}
	\item $\mathbf{b} = \mathbf{W}\mathbf{y}$ is computed according to lemma \ref{thm:matrix_const};
	\item $\mathbf{e} = \mathbf{S}\boldsymbol\beta_{l-1}$ is computed according to lemma \ref{thm:matrix_vector};
	\item $\mathbf{c} = \mathbf{b} + \mathbf{e}$ is computed according to lemma \ref{thm:sum_vectors};
	\item $\boldsymbol\beta_{l} = h_\lambda(\mathbf{c})$ is computed according to lemma \ref{thm:soft_thresholding}.
\end{enumerate}
\end{theorem}

\subsection{Approximation quality}
\label{sec:approx_quality}
We have used two approximations in forward propagation of the uncertainty. First, in lemma \ref{thm:matrix_vector} a Gaussian matrix is multiplied by a spike-and-slab vector and their product is approximated with the Gaussian distribution. Second, in lemma \ref{thm:soft_thresholding} the result of soft-thresholding of a Gaussian vector is approximated with the spike-and-slab distribution. In this section we demonstrate that these approximations are close to the real distributions.
